---
title: "NSCLC_Modeling_3"
format: pdf
editor: visual
---

(Number of altered genes in the pathway for the patient) / (Total number of genes in that KEGG pathway).

## Setups

```{r}
# install.packages("msigdbr"): this is to pull the KEGG pathway
library(dplyr)
library(tibble)
library(msigdbr)
library(timeROC)
library(survminer)
library(tidyr)
library(glmnet)
library(LTRCforests)
library(ggplot2)
library(survival)
nsclc <- read.csv("nsclc.csv") # read in the dataset

# Adjusting factor variables
nsclc <- nsclc %>% mutate(
  stage_f = as.factor(stage_f),
  smoke_f = as.factor(smoke_f),
  prior_med_f = as.factor(prior_med_to_msk_f),
  gender = ifelse(GENDER == "Female", 1 ,0) # change Female to 1, and male to 0
) %>% select(-c(
  "GENDER", "STAGE_CDM_DERIVED", 
  "SMOKING_PREDICTIONS_3_CLASSES", "PRIOR_MED_TO_MSK"
))

# Define clinical columns to isolate the gene matrix

clinical_cols <- c(

  "X", "PATIENT_ID", "age_at_diagnosis",  "event", 

  "entry_time", "exit_time", "smoke_f", "stage_f", "prior_med_to_msk_f", 

  "prior_med_f", "gender"

)

# Identify genes by subtracting clinical columns from all column names
gene_cols <- setdiff(names(nsclc), clinical_cols)

# Create the matrix used in the loop
gene_matrix <- nsclc[, gene_cols]
```

## Data Processing

```{r}

# Pull KEGG pathways from MSigDB
# C2 category, CP:KEGG subcategory contains all KEGG pathways
kegg_data <- msigdbr(
  species = "Homo sapiens", 
  collection = "C2", 
  subcollection = "CP:KEGG_LEGACY"
)
#  Convert to a list where each element is a pathway containing its gene symbols
kegg_list <- split(x = kegg_data$gene_symbol, f = kegg_data$gs_name)

#  Create an empty list to store the fractional scores for each pathway
pathway_scores <- list()

# Loop through each KEGG pathway to calculate the fraction
for (pathway_name in names(kegg_list)) {
  
  # Get the unique genes in this specific KEGG pathway
  pathway_genes <- unique(kegg_list[[pathway_name]])
  total_genes_in_pathway <- length(pathway_genes)
  
  intersecting_genes <- intersect(pathway_genes, gene_cols)
    if (length(intersecting_genes) > 0) {
    patient_mutations <- rowSums(gene_matrix[, intersecting_genes, drop = FALSE], na.rm = TRUE)
    fraction <- patient_mutations / total_genes_in_pathway
    pathway_scores[[pathway_name]] <- fraction
  }
}

#Combine the list into a dataframe
pathway_matrix <- bind_cols(pathway_scores)

# Filter out pathways with zero variance
pathway_df <- colSums(pathway_matrix, na.rm = TRUE)
active_pathways <- names(pathway_df[pathway_df > 0])
pathway_matrix_filtered <- pathway_matrix[, active_pathways]

# Recombine 
nsclc_pathway <- bind_cols(nsclc[, clinical_cols], pathway_matrix_filtered)

cat("Total KEGG Pathways evaluated:", length(kegg_list), "\n")
cat("Pathways with at least 1 mutation in cohort:", length(active_pathways), "\n")
```

2.  Prepare data for glmnet

```{r}
library(tidyr)

# Create a complete dataset 
nsclc_path_complete <- nsclc_pathway %>% drop_na()

# Isolate the predictor variables 
x_vars_path <- nsclc_path_complete %>% 
  select(-X, -PATIENT_ID, -entry_time, -exit_time, -event)

# Create the predictor matrix 
x_matrix_path <- model.matrix(~ . - 1, data = x_vars_path)

# Create the Survival response object
y_surv_path <- Surv(
  time = nsclc_path_complete$entry_time, 
  time2 = nsclc_path_complete$exit_time, 
  event = nsclc_path_complete$event
)
```

## 1. Cross-validation with Lasso

```{r}
set.seed(629)

# Fit the Cross-Validated Lasso Cox Model using Pathway features
cv_lasso_path <- cv.glmnet(
  x = x_matrix_path,
  y = y_surv_path,
  family = "cox",
  alpha = 1,          
  type.measure = "C"  
)

plot(cv_lasso_path)
title("LASSO Penalty on KEGG Pathways", line = 2.5)

cat("Optimal Lambda:", cv_lasso_path$lambda.min, "\n")
cat("Max CV C-index (Pathways):", max(cv_lasso_path$cvm, na.rm = TRUE), "\n")

# Extract the winning pathway features
lasso_coefs_path <- coef(cv_lasso_path, s = "lambda.min")
active_features_path <- rownames(lasso_coefs_path)[which(lasso_coefs_path != 0)]

cat("Features selected by LASSO:", length(active_features_path), "\n\n")

# View the selected features and weights
selected_weights_path <- lasso_coefs_path[active_features_path, , drop = FALSE]
print(selected_weights_path)
```

### Post-lasso Unpenalized Cox Model

```{r}
# Identify ONLY the pathways that LASSO selected
selected_pathways <- active_features_path[active_features_path %in% colnames(pathway_matrix_filtered)]

# Check if any selected pathways have essentially 0 variance in the COMPLETE dataset
# If a pathway has only 0s, scale() will fail and coxph will crash.
pathway_variances <- apply(nsclc_path_complete[, selected_pathways, drop=FALSE], 2, var)
valid_pathways <- names(pathway_variances[pathway_variances > 0])

if(length(valid_pathways) < length(selected_pathways)) {
  cat("Dropped", length(selected_pathways) - length(valid_pathways), 
      "pathways due to zero variance in the clean dataset.\n")
}

# Build final data frame
final_model_data_path <- nsclc_path_complete %>%
  select(
    entry_time, exit_time, event,   
    age_at_diagnosis, smoke_f, stage_f, prior_med_f, gender, 
    all_of(valid_pathways)         
  ) %>%
  mutate(across(all_of(valid_pathways), ~ as.numeric(scale(.))))

# Run the final Cox model
clean_cox_model_path <- coxph(
  Surv(entry_time, exit_time, event) ~ .,
  data = final_model_data_path,
  control = coxph.control(iter.max = 100) 
)

# Check the output  
summary(clean_cox_model_path)
```

### C-index

```{r}
# Setup CV
set.seed(629)
K <- 5
n_path <- nrow(final_model_data_path)
fold_id_path <- sample(rep(1:K, length.out = n_path))
cv_cindex_post_lasso_path <- numeric(K)

for(k in 1:K){
  
  train_data_path <- final_model_data_path[fold_id_path != k, ]
  test_data_path  <- final_model_data_path[fold_id_path == k, ]
  
  # Fit the post-LASSO model on the training fold 
  # (Using iter.max = 100 to prevent convergence warnings on smaller folds)
  fit_path <- coxph(
    Surv(entry_time, exit_time, event) ~ .,
    data = train_data_path,
    control = coxph.control(iter.max = 100)
  )
  
  # Predict risk scores on the test fold
  test_data_path$risk_score <- predict(fit_path, newdata = test_data_path, type = "lp", na.action = na.pass)
  
  # Compute C-index on the test fold
  c_obj_path <- concordance(
    Surv(entry_time, exit_time, event) ~ risk_score,
    data = test_data_path,
    reverse = TRUE
  )
  
  cv_cindex_post_lasso_path[k] <- c_obj_path$concordance
}

# Print results
cat("Pathway Model Mean CV C-index:", mean(cv_cindex_post_lasso_path, na.rm = TRUE), "\n")
cat("Pathway Model SD CV C-index:", sd(cv_cindex_post_lasso_path, na.rm = TRUE), "\n")
```

### Time-Dependent ROC Curve

```{r}
 
# Generate the risk scores (linear predictors) from your final pathway model
final_model_data_path$risk_score <- predict(clean_cox_model_path, type = "lp")

# Define the time horizon (80th percentile of follow-up time)
time_horizon_path <- quantile(final_model_data_path$exit_time, 0.80)

#  Calculate the Time-dependent ROC
roc_pathway <- timeROC(
  T = final_model_data_path$exit_time,
  delta = final_model_data_path$event,
  marker = final_model_data_path$risk_score, 
  cause = 1,
  times = time_horizon_path,
  iid = TRUE  
)

# Extract the Area Under the Curve (AUC) and Confidence Intervals
auc_value_path <- roc_pathway$AUC[2] # Index 2 corresponds to time_horizon_path
se_value_path <- roc_pathway$inference$vect_sd_1[2]

# Calculate 95% CI
ci_lower_path <- auc_value_path - 1.96 * se_value_path
ci_upper_path <- auc_value_path + 1.96 * se_value_path

cat("Time Horizon:", round(time_horizon_path, 2), "\n")
cat("Pathway AUC at Time Horizon:", round(auc_value_path, 3), "\n")
cat("95% CI: [", round(ci_lower_path, 3), "-", round(ci_upper_path, 3), "]\n\n")

# Plot the ROC Curve
{plot(roc_pathway, 
     time = time_horizon_path, 
     col = "darkblue", # Changed color to visually distinguish from your gene model
     lwd = 2, 
     title = FALSE)

# Add a reference line and a title
abline(0, 1, lty = 2, col = "gray50")
title(main = paste("Pathway Cox Model ROC\nat Time =", round(time_horizon_path, 1)))

# Add a legend with the AUC value
legend("bottomright", 
       legend = paste0("AUC = ", round(auc_value_path, 2), 
                       " (", round(ci_lower_path, 2), "-", round(ci_upper_path, 2), ")"),
       col = "darkblue", 
       lwd = 2, 
       bty = "n")}
```

### calibration plot

```{r}
 
#  Define the time horizon (80th percentile of follow-up time, same as ROC)
time_horizon_path <- quantile(final_model_data_path$exit_time, 0.80)

# Calculate predicted survival probabilities for the entire cohort at the time horizon
sf_final_path <- survfit(clean_cox_model_path, newdata = final_model_data_path)
pred_surv_path <- summary(sf_final_path, times = time_horizon_path)$surv
pred_surv_path <- as.numeric(pred_surv_path)

# Add predictions to the dataset
final_model_data_path$pred_surv <- pred_surv_path

# Create 10 risk groups (deciles) based on predicted survival
final_model_data_path$decile <- ntile(final_model_data_path$pred_surv, 10)

# Calculate observed Kaplan-Meier estimates for each decile
calibration_data_path <- final_model_data_path %>%
  group_by(decile) %>%
  group_modify(~ {
    
    km_fit <- survfit(Surv(entry_time, exit_time, event) ~ 1, data = .x)
 
    km_summary <- summary(km_fit, 
                          times = time_horizon_path, 
                          extend = TRUE)
    
    data.frame(
      mean_pred = mean(.x$pred_surv, na.rm = TRUE),
      km_est = km_summary$surv
    )
  })

# Plot the calibration curve
ggplot(calibration_data_path, aes(x = mean_pred, y = km_est)) +
  geom_point(size = 3, color = "darkblue") +
  geom_line(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = "Predicted Survival Probability",
    y = "Observed Survival Probability (Kaplan-Meier)",
    title = "Calibration Plot: Pathway-Based Post-LASSO Cox Model",
    subtitle = paste("Evaluated at Time =", round(time_horizon_path, 1), "| Red dashed line = perfect calibration")
  ) +
  theme_minimal()
```

## 2. Left-truncated random forest

### Data Cleaning and Train/Test Split (80/20)

```{r}
# Clean the dataset to ensure the forest only sees true predictors
rsf_data <- final_model_data_path %>%
  select(-any_of(c("risk_score", "pred_surv", "decile"))) %>%
  as.data.frame()

set.seed(629)
N_total <- nrow(rsf_data)
train_idx <- sample(1:N_total, size = floor(0.8 * N_total))

train_data_rsf <- rsf_data[train_idx, ]
test_data_rsf  <- rsf_data[-train_idx, ]

cat("Train patients:", nrow(train_data_rsf), "\n")
cat("Test patients:", nrow(test_data_rsf), "\n")
```

### Fit the LTRC Forest on TRAIN Data

```{r}
predictor_names <- setdiff(colnames(rsf_data), c("entry_time", "exit_time", "event"))

# Build the formula dynamically
rsf_formula <- as.formula(
  paste("Surv(entry_time, exit_time, event) ~", paste(predictor_names, collapse = " + "))
)

# Fit the forest using the training data
set.seed(629)
final_rsf_model <- ltrcrrf(
  formula = rsf_formula,
  data = train_data_rsf,
  ntree = 500,        
  nodesize = 20,      
  mtry = 3,           
  nsplit = 10         
)
```

### Evaluation on Independent TEST Data

#### c-index

```{r}
# Define the time horizon based on the training set follow-up
time_horizon_path <- quantile(train_data_rsf$exit_time, 0.80)

# Predict survival probabilities on the TEST set
test_pred_obj <- predictProb(
  final_rsf_model, 
  newdata = test_data_rsf, 
  time.eval = time_horizon_path
)

# Add predictions and risk scores to the test dataset
test_data_rsf$pred_surv <- as.numeric(test_pred_obj$survival.probs)
test_data_rsf$rsf_risk <- 1 - test_data_rsf$pred_surv

c_obj_rsf <- concordance(
  Surv(entry_time, exit_time, event) ~ rsf_risk,
  data = test_data_rsf,
  reverse = TRUE
)

cat("\nOut-of-Sample C-index:", round(c_obj_rsf$concordance, 3), "\n")
```


#### Time-Dependent ROC on TEST

```{r}
roc_rsf <- timeROC(
  T = test_data_rsf$exit_time,
  delta = test_data_rsf$event,
  marker = test_data_rsf$rsf_risk, 
  cause = 1,
  times = time_horizon_path,
  iid = TRUE
)

auc_value_rsf <- roc_rsf$AUC[2] 
se_value_rsf <- roc_rsf$inference$vect_sd_1[2]
ci_lower_rsf <- auc_value_rsf - 1.96 * se_value_rsf
ci_upper_rsf <- auc_value_rsf + 1.96 * se_value_rsf

cat("Out-of-Sample AUC at Time Horizon (", round(time_horizon_path, 1), "):", round(auc_value_rsf, 3), "\n")
cat("95% CI: [", round(ci_lower_rsf, 3), "-", round(ci_upper_rsf, 3), "]\n\n")

plot(roc_rsf, 
     time = time_horizon_path, 
     col = "navy", 
     lwd = 2, 
     title = FALSE)

abline(0, 1, lty = 2, col = "gray")
title(main = paste("Out-of-Sample LTRC Forest ROC\nat Time =", round(time_horizon_path, 1)))

legend("bottomright", 
       legend = paste0("AUC = ", round(auc_value_rsf, 2), 
                       " (", round(ci_lower_rsf, 2), "-", round(ci_upper_rsf, 2), ")"),
       col = "navy", 
       lwd = 2, 
       bty = "n")
```

#### Calibration Plot on TEST

```{r}
# Using 5 groups (quintiles) instead of 10 deciles because the test set is 20% of the total size.
test_data_rsf$decile <- ntile(test_data_rsf$pred_surv, 5)

calibration_data_rsf <- test_data_rsf %>%
  group_by(decile) %>%
  group_modify(~ {
    
    km_fit <- survfit(Surv(entry_time, exit_time, event) ~ 1, data = .x)
    km_summary <- summary(km_fit, 
                          times = time_horizon_path, 
                          extend = TRUE)
    
    data.frame(
      mean_pred = mean(.x$pred_surv, na.rm = TRUE),
      km_est = km_summary$surv
    )
  })

ggplot(calibration_data_rsf, aes(x = mean_pred, y = km_est)) +
  geom_point(size = 3, color = "navy") +
  geom_line(color = "navy") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = "Predicted Survival Probability",
    y = "Observed Survival Probability (Kaplan-Meier)",
    title = "Out-of-Sample Calibration: LTRC Forest"
  ) +
  theme_minimal()
```
