---
title: "NSCLC_Modeling"
format: pdf
editor: visual
---

## Setups

```{r}
library(dplyr)
library(survminer)
library(ggplot2)
library(timeROC)
library(LTRCforests)
library(tidyr)
library(LTRCtrees)
library(survival)
nsclc <- read.csv("nsclc.csv") # read in the dataset

# Adjusting factor variables
nsclc <- nsclc %>% mutate(
  stage_f = as.factor(stage_f),
  smoke_f = as.factor(smoke_f),
  prior_med_f = as.factor(prior_med_to_msk_f),
  gender = ifelse(GENDER == "Female", 1 ,0) # change Female to 1, and male to 0
) %>% select(-c(
  "GENDER", "STAGE_CDM_DERIVED", 
  "SMOKING_PREDICTIONS_3_CLASSES", "PRIOR_MED_TO_MSK"
))

```

## 1. Cox - Frequency Filtering: 

1. Drop any gene mutation that occurs in less than < 3% of the cohort.

```{r}
# Define threshold
threshold <- 0.03

# clinical/demographics columns to be kept
clinical_cols <- c(
  "X", "PATIENT_ID", "age_at_diagnosis",  "event", 
  "entry_time", "exit_time", "smoke_f", "stage_f", "prior_med_to_msk_f", 
  "prior_med_f", "gender"
)

# select all gene columns
gene_cols <- setdiff(colnames(nsclc), clinical_cols)

# calculate the mutation frequency for all genes
gene_frequencies <- colMeans(nsclc[, gene_cols], na.rm = TRUE)

# identify genes meet or exceed the threshold
genes_to_keep <- names(gene_frequencies[gene_frequencies >= threshold])

# dataset with the clinical columns + the filtered genes
final_cols <- c(clinical_cols, genes_to_keep)
nsclc_filtered <- nsclc[, final_cols]

# check how many genes were dropped vs. kept
cat("Original number of genes:", length(gene_cols), "\n")
cat("Genes kept (>= 3%):", length(genes_to_keep), "\n")
cat("Genes dropped:", length(gene_cols) - length(genes_to_keep), "\n")
```

### cox ph model:

```{r}
library(tidyr)
library(glmnet)

# Drop missing values so x and y matrices align perfectly
nsclc_filtered_complete <- nsclc_filtered %>% drop_na()

# Create the predictor matrix and survival object
x_vars_freq <- nsclc_filtered_complete %>% 
  select(-X, -PATIENT_ID, -entry_time, -exit_time, -event)
x_matrix_freq <- model.matrix(~ . - 1, data = x_vars_freq)

y_surv_freq <- Surv(
  time = nsclc_filtered_complete$entry_time, 
  time2 = nsclc_filtered_complete$exit_time, 
  event = nsclc_filtered_complete$event
)

# Run Cross-Validated LASSO
set.seed(629)
cv_lasso_freq <- cv.glmnet(
  x = x_matrix_freq, 
  y = y_surv_freq, 
  family = "cox", 
  alpha = 1, 
  type.measure = "C"
)
```


### Extract winning features and isolate just the genes
```{r}
lasso_coefs_freq <- coef(cv_lasso_freq, s = "lambda.min")
active_features_freq <- rownames(lasso_coefs_freq)[which(lasso_coefs_freq != 0)]
selected_genes_freq <- active_features_freq[active_features_freq %in% genes_to_keep]

cat("Frequency Filter + LASSO selected", length(selected_genes_freq), "genes.\n")
```

### Post-LASSO Cox Model
```{r}
# Build clean dataset avoiding the dummy-variable trap for clinical factors
final_data_freq <- nsclc_filtered_complete %>%
  select(
    entry_time, exit_time, event, 
    age_at_diagnosis, smoke_f, stage_f, prior_med_f, gender, 
    all_of(selected_genes_freq)
  )

clean_cox_freq <- coxph(
  Surv(entry_time, exit_time, event) ~ .,
  data = final_data_freq
)

summary(clean_cox_freq)
```

### cross validation

```{r}
### 4. Cross-Validation (C-Index)
set.seed(629)
K <- 5
n <- nrow(final_data_freq)
fold_id <- sample(rep(1:K, length.out = n))
cv_cindex_freq <- numeric(K)

for(k in 1:K){
  train_data <- final_data_freq[fold_id != k, ]
  test_data  <- final_data_freq[fold_id == k, ]
  
  fit <- coxph(Surv(entry_time, exit_time, event) ~ ., data = train_data)
  
  test_data$risk_score <- predict(fit, newdata = test_data, type = "lp", na.action = na.pass)
  
  c_obj <- concordance(
    Surv(entry_time, exit_time, event) ~ risk_score,
    data = test_data,
    reverse = TRUE
  )
  
  cv_cindex_freq[k] <- c_obj$concordance
}

cat("Frequency Pipeline Mean C-index:", mean(cv_cindex_freq, na.rm = TRUE), "\n")
cat("Frequency Pipeline SD C-index:", sd(cv_cindex_freq, na.rm = TRUE), "\n")
```

### calibration visualization

```{r}
library(ggplot2)
library(dplyr)
library(survival)

# 1. Define the time horizon (80th percentile of follow-up time)
# Make sure to use the exit_time from your frequency dataset
time_horizon_freq <- quantile(final_data_freq$exit_time, 0.80)

# 2. Calculate predicted survival probabilities at the time horizon
sf_final_freq <- survfit(clean_cox_freq, newdata = final_data_freq)
pred_surv_freq <- summary(sf_final_freq, times = time_horizon_freq)$surv
pred_surv_freq <- as.numeric(pred_surv_freq)

# 3. Add predictions to the dataset and create risk deciles
final_data_freq$pred_surv <- pred_surv_freq
final_data_freq$decile <- ntile(final_data_freq$pred_surv, 10)

# 4. Calculate observed Kaplan-Meier estimates for each decile
calibration_data_freq <- final_data_freq %>%
  group_by(decile) %>%
  group_modify(~ {
    
    km_fit <- survfit(Surv(entry_time, exit_time, event) ~ 1, data = .x)
    
    # extend = TRUE ensures we get an estimate even if the curve drops off early
    km_summary <- summary(km_fit, 
                          times = time_horizon_freq, 
                          extend = TRUE)
    
    data.frame(
      mean_pred = mean(.x$pred_surv, na.rm = TRUE),
      km_est = km_summary$surv
    )
  })

# 5. Plot the calibration curve
ggplot(calibration_data_freq, aes(x = mean_pred, y = km_est)) +
  geom_point(size = 3, color = "darkblue") + # Changed color to easily distinguish from TF-IDF
  geom_line(color = "darkblue", size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = paste("Predicted Survival Probability at Time =", round(time_horizon_freq, 1)),
    y = "Observed Survival Probability (Kaplan-Meier)",
    title = "Calibration Plot: Frequency Filter (>3%) Model",
    subtitle = "Red dashed line indicates perfect calibration"
  ) +
  theme_minimal()
```

```{r}
# Make sure the package is loaded

# 1. Generate the risk scores from your frequency-based model
final_data_freq$risk_score <- predict(clean_cox_freq, type = "lp")

# 2. Define the time horizon (ensuring it matches your calibration plot)
time_horizon_freq <- quantile(final_data_freq$exit_time, 0.80)

# 3. Calculate the Time-dependent ROC
roc_freq <- timeROC(
  T = final_data_freq$exit_time,
  delta = final_data_freq$event,
  marker = final_data_freq$risk_score, 
  cause = 1,
  times = time_horizon_freq,
  iid = TRUE # Keeps the data needed to calculate Confidence Intervals
)

# 4. Extract the Area Under the Curve (AUC) and 95% Confidence Intervals
auc_val_freq <- roc_freq$AUC[2] 
se_val_freq <- roc_freq$inference$vect_sd_1[2]

ci_lower_freq <- auc_val_freq - 1.96 * se_val_freq
ci_upper_freq <- auc_val_freq + 1.96 * se_val_freq

cat("Frequency Pipeline - Time Horizon:", round(time_horizon_freq, 2), "\n")
cat("Frequency Pipeline - AUC:", round(auc_val_freq, 3), "\n")
cat("95% CI: [", round(ci_lower_freq, 3), "-", round(ci_upper_freq, 3), "]\n\n")

# 5. Plot the ROC Curve
# Using 'dodgerblue' to match your Frequency calibration plot
plot(roc_freq, 
     time = time_horizon_freq, 
     col = "darkblue", 
     lwd = 2, 
     title = FALSE)

# Add reference line and title
abline(0, 1, lty = 2, col = "gray50")
title(main = paste("Frequency Filter (>3%) Model ROC\nat Time =", round(time_horizon_freq, 1)))

# Add legend with the final AUC value
legend("bottomright", 
       legend = paste0("AUC = ", round(auc_val_freq, 2), 
                       " (", round(ci_lower_freq, 2), "-", round(ci_upper_freq, 2), ")"),
       col = "darkblue", 
       lwd = 2, 
       bty = "n")
```

## 2. RSF

### Train/Test Split (80/20)

```{r}
set.seed(629)
N_total <- nrow(nsclc_filtered_complete)
train_idx <- sample(1:N_total, size = floor(0.8 * N_total))

train_data_rf <- nsclc_filtered_complete[train_idx, ]
test_data_rf  <- nsclc_filtered_complete[-train_idx, ]

cat("Train patients:", nrow(train_data_rf), "\n")
cat("Test patients:", nrow(test_data_rf), "\n")
```

### Fit the RSF on TRAIN Data

```{r}
# Define the features to include in the model
clinical_features <- c("age_at_diagnosis", "smoke_f", "stage_f", "prior_med_f", "gender")
all_features <- c(clinical_features, genes_to_keep)

# Construct the formula
rf_formula <- as.formula(
  paste("Surv(entry_time, exit_time, event) ~", paste(all_features, collapse = " + "))
)

# Define mtry
m_try <- max(floor(sqrt(length(all_features))), 1)

# Fit the Random Forest Model ONLY on the training data
set.seed(629)
rf_model <- ltrcrrf(
  formula = rf_formula,
  data = train_data_rf,
  ntree = 500,        
  nodesize = 20,      
  mtry = m_try,       
  nsplit = 10         
)
```

### Evaluation on Independent TEST Data

#### c-index

```{r}
# Define the time horizon based on the training set follow-up
time_horizon_rf <- quantile(train_data_rf$exit_time, 0.80)

# Predict survival probabilities on the TEST set
test_prob_obj <- predictProb(
  rf_model,
  newdata = test_data_rf,
  time.eval = time_horizon_rf
)

# Extract survival and calculate risk on the test set
test_data_rf$pred_surv <- as.numeric(test_prob_obj$survival.probs)
test_data_rf$risk_score <- 1 - test_data_rf$pred_surv


# c-index
c_obj <- concordance(
  Surv(entry_time, exit_time, event) ~ risk_score,
  data = test_data_rf,
  reverse = TRUE 
)

cat("\nOut-of-Sample C-index:", round(c_obj$concordance, 3), "\n")
```

#### Time-Dependent ROC on TEST

```{r}
roc_rf <- timeROC(
  T = test_data_rf$exit_time,
  delta = test_data_rf$event,
  marker = test_data_rf$risk_score, 
  cause = 1,
  times = time_horizon_rf,
  iid = TRUE 
)

auc_val_rf <- roc_rf$AUC[2] 
se_val_rf <- roc_rf$inference$vect_sd_1[2]
ci_lower_rf <- auc_val_rf - 1.96 * se_val_rf
ci_upper_rf <- auc_val_rf + 1.96 * se_val_rf

cat("Out-of-Sample AUC at Time Horizon (", round(time_horizon_rf, 1), "):", round(auc_val_rf, 3), "\n")
cat("95% CI: [", round(ci_lower_rf, 3), "-", round(ci_upper_rf, 3), "]\n\n")

# Plot the ROC Curve
plot(roc_rf, 
     time = time_horizon_rf, 
     col = "navy", 
     lwd = 2, 
     title = FALSE)

abline(0, 1, lty = 2, col = "gray")
title(main = paste("Out-of-Sample Baseline RSF ROC\nat Time =", round(time_horizon_rf, 1)))
legend("bottomright", 
       legend = paste0("AUC = ", round(auc_val_rf, 2), 
                       " (", round(ci_lower_rf, 2), "-", round(ci_upper_rf, 2), ")"),
       col = "navy", 
       lwd = 2, 
       bty = "n")
```

#### Calibration Plot on TEST

```{r}
# Using 5 bins (quintiles) due to the reduced size of the 20% test dataset
test_data_rf$decile <- ntile(test_data_rf$pred_surv, 5)

calibration_data_rf <- test_data_rf %>%
  group_by(decile) %>%
  group_modify(~ {
    km_fit <- survfit(Surv(entry_time, exit_time, event) ~ 1, data = .x)
    km_summary <- summary(km_fit, times = time_horizon_rf, extend = TRUE)
    
    data.frame(
      mean_pred = mean(.x$pred_surv, na.rm = TRUE),
      km_est = km_summary$surv
    )
  })

ggplot(calibration_data_rf, aes(x = mean_pred, y = km_est)) +
  geom_point(size = 3, color = "navy") +
  geom_line(color = "navy") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = paste("Predicted Survival Probability at Time =", round(time_horizon_rf, 1)),
    y = "Observed Survival Probability (Kaplan-Meier)",
    title = "Out-of-Sample Calibration: Baseline RSF"
  ) +
  theme_minimal()
```

