---
title: "NSCLC_Modeling_2"
format: pdf
editor: visual
---

## Setups

```{r}
library(dplyr)
library(survminer)
library(LTRCforests)
library(ggplot2)
library(survival)
nsclc <- read.csv("nsclc.csv") # read in the dataset

# Adjusting factor variables
# Adjusting factor variables
nsclc <- nsclc %>% mutate(
  stage_f = as.factor(stage_f),
  smoke_f = as.factor(smoke_f),
  prior_med_f = as.factor(prior_med_to_msk_f),
  gender = ifelse(GENDER == "Female", 1 ,0) # change Female to 1, and male to 0
) %>% select(-c(
  "GENDER", "STAGE_CDM_DERIVED", 
  "SMOKING_PREDICTIONS_3_CLASSES", "PRIOR_MED_TO_MSK"
))
```

## 1. Cox - Rarity weight

```{r}
# 1. Define clinical columns to isolate the gene matrix
clinical_cols <- c(
  "X", "PATIENT_ID", "age_at_diagnosis",  "event", 
  "entry_time", "exit_time", "smoke_f", "stage_f", "prior_med_to_msk_f", 
  "prior_med_f", "gender"
)

# Isolate the true gene columns
gene_cols <- setdiff(colnames(nsclc), clinical_cols)
gene_matrix <- nsclc %>%
  select(all_of(gene_cols)) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.)))) 

# 2. Calculate N and df 
N <- nrow(gene_matrix)
df <- colSums(gene_matrix, na.rm = TRUE) 

# 3. Handle zero-variance genes
# If a gene has 0 mutations, log(N/0) equals Infinity. We must drop them.
# It is also highly recommended to drop "singletons" (genes mutated in only 1 or 2 patients) 
# to prevent the Cox model from overfitting to a single patient's outcome.
min_mutations <- 3 
genes_to_keep_tfidf <- names(df[df >= min_mutations])

# Subset matrix and frequencies to kept genes
gene_matrix_filtered <- gene_matrix[, genes_to_keep_tfidf]
df_filtered <- df[df >= min_mutations]

# 4. Calculate the IDF Weights
idf_weights <- log(N / df_filtered)

# 5. Apply weights (TF * IDF)
# sweep() multiplies each column in the matrix by its corresponding IDF weight
# If patient = 1 (mutated), they get 1 * IDF_weight. If 0, they get 0.
gene_matrix_weighted <- sweep(
  x = gene_matrix_filtered, 
  MARGIN = 2, 
  STATS = idf_weights, 
  FUN = "*"
)

# 6. Recombine with clinical data
nsclc_tfidf <- bind_cols(nsclc[, clinical_cols], gene_matrix_weighted)

# Check your results
cat("Total patients (N):", N, "\n")
cat("Genes evaluated:", length(gene_cols), "\n")
cat("Genes kept for TF-IDF (>= 3 mutations):", length(genes_to_keep_tfidf), "\n")
```

### Data Preparation for glmnet

```{r}
library(glmnet)

# 1. Isolate the predictor variables
x_vars <- nsclc_tfidf %>% 
  select(-X, -PATIENT_ID, -entry_time, -exit_time, -event)

# 2. Create the predictor matrix
# The '~ . - 1' tells R to create dummy variables for your factors but omit the intercept
# (Cox models don't use a standard intercept in the x matrix)
x_matrix <- model.matrix(~ . - 1, data = x_vars)

# 3. Create the Survival response object
# glmnet accepts standard Surv() objects for the 'cox' family
y_surv <- Surv(
  time = nsclc_tfidf$entry_time, 
  time2 = nsclc_tfidf$exit_time, 
  event = nsclc_tfidf$event
)
```

### data cleaning

```{r}
library(tidyr)

# 1. Create a complete dataset with NO missing values
# This guarantees x and y will be perfectly aligned
nsclc_complete <- nsclc_tfidf %>% drop_na()

# 2. Isolate the predictor variables from the COMPLETE dataset
x_vars <- nsclc_complete %>% 
  select(-X, -PATIENT_ID, -entry_time, -exit_time, -event)

# 3. Create the predictor matrix
x_matrix <- model.matrix(~ . - 1, data = x_vars)

# 4. Create the Survival response object from the SAME complete dataset
y_surv <- Surv(
  time = nsclc_complete$entry_time, 
  time2 = nsclc_complete$exit_time, 
  event = nsclc_complete$event
)
```


### fit cross-validated lasso model

```{r}
# Set seed for reproducibility in cross-validation
set.seed(629)

# Fit the Cross-Validated Lasso Cox Model
cv_lasso_cox <- cv.glmnet(
  x = x_matrix,
  y = y_surv,
  family = "cox",
  alpha = 1,          # alpha = 1 specifies LASSO (alpha = 0 would be Ridge)
  type.measure = "C"  # Optimize the penalty to maximize the Harrell's C-index
)

# Plot the cross-validation curve
# This shows the C-index on the y-axis against the log(lambda) penalty on the x-axis
plot(cv_lasso_cox)

# Print the performance metrics
cat("Optimal Lambda (maximizes C-index):", cv_lasso_cox$lambda.min, "\n")
cat("Max Cross-Validated C-index:", max(cv_lasso_cox$cvm, na.rm = TRUE), "\n")
```
### Extract the "Winning" Features


```{r}
# Extract the coefficients at the optimal lambda value
lasso_coefs <- coef(cv_lasso_cox, s = "lambda.min")

# Filter for features that have a non-zero coefficient
active_features <- rownames(lasso_coefs)[which(lasso_coefs != 0)]

cat("Total features evaluated:", ncol(x_matrix), "\n")
cat("Features selected by LASSO:", length(active_features), "\n\n")

# View the final selected features and their weights
selected_weights <- lasso_coefs[active_features, , drop = FALSE]
print(selected_weights)
```
### Post-LASSO unpenalized Cox model.

```{r}
# 1. Identify ONLY the genes that LASSO selected
# We do this by cross-referencing the active features with your original gene list
selected_genes <- active_features[active_features %in% gene_cols]

# 2. Build a new dataframe using the ORIGINAL clinical columns + the winning genes
# This brings back your proper factor variables (stage_f, prior_med_f, etc.)
final_model_data_clean <- nsclc_complete %>%
  select(
    entry_time, exit_time, event,   # Survival outcomes
    age_at_diagnosis, smoke_f, stage_f, prior_med_f, gender, # Clinical vars
    all_of(selected_genes)          # Winning genes
  )

# 3. Run the final Cox model
# R will automatically handle the dummy coding behind the scenes
clean_cox_model <- coxph(
  Surv(entry_time, exit_time, event) ~ .,
  data = final_model_data_clean
)

# Check the beautiful, NA-free output
summary(clean_cox_model)
```


### c-index

```{r}

# Setup CV
set.seed(629)
K <- 5
n <- nrow(final_model_data_clean)
fold_id <- sample(rep(1:K, length.out = n))
cv_cindex_post_lasso <- numeric(K)

for(k in 1:K){
  
  train_data <- final_model_data_clean[fold_id != k, ]
  test_data  <- final_model_data_clean[fold_id == k, ]
  
  # Fit the post-LASSO model on the training fold 
  fit <- coxph(
    Surv(entry_time, exit_time, event) ~ .,
    data = train_data
  )
  
  # Predict risk scores on the test fold
  test_data$risk_score <- predict(fit, newdata = test_data, type = "lp", na.action = na.pass)
  
  # Compute C-index on the test fold
  c_obj <- concordance(
    Surv(entry_time, exit_time, event) ~ risk_score,
    data = test_data,
    reverse = TRUE
  )
  
  cv_cindex_post_lasso[k] <- c_obj$concordance
}

# Print results
cat("Post-LASSO Mean C-index:", mean(cv_cindex_post_lasso, na.rm = TRUE), "\n")
cat("Post-LASSO SD C-index:", sd(cv_cindex_post_lasso, na.rm = TRUE), "\n")
```



### calibration plot:

```{r}
library(ggplot2)

# Define the time horizon (e.g., 80th percentile of follow-up time)
time_horizon <- quantile(final_model_data_clean$exit_time, 0.80)

# Calculate predicted survival probabilities for the entire cohort at the time horizon
sf_final <- survfit(clean_cox_model, newdata = final_model_data_clean)
pred_surv <- summary(sf_final, times = time_horizon)$surv
pred_surv <- as.numeric(pred_surv)

# Add predictions to the dataset
final_model_data_clean$pred_surv <- pred_surv

# Create 10 risk groups (deciles) based on predicted survival
final_model_data_clean$decile <- ntile(final_model_data_clean$pred_surv, 10)

# Calculate observed Kaplan-Meier estimates for each decile
calibration_data <- final_model_data_clean %>%
  group_by(decile) %>%
  group_modify(~ {
    
    km_fit <- survfit(Surv(entry_time, exit_time, event) ~ 1, data = .x)
    
    km_summary <- summary(km_fit, 
                          times = time_horizon, 
                          extend = TRUE)
    
    data.frame(
      mean_pred = mean(.x$pred_surv, na.rm = TRUE),
      km_est = km_summary$surv
    )
  })

# Plot the calibration curve
ggplot(calibration_data, aes(x = mean_pred, y = km_est)) +
  geom_point(size = 3, color = "darkblue") +
  geom_line(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = "Predicted Survival Probability",
    y = "Observed Survival Probability (Kaplan-Meier)",
    title = "Calibration Plot: Post-LASSO Cox Model",
    subtitle = "Red dashed line indicates perfect calibration"
  ) +
  theme_minimal()
```

### Time-Dependent ROC Curve

```{r}
library(timeROC)

# 1. Generate the risk scores (linear predictors) from your final model
# We use the clean model and dataset from the previous step
final_model_data_clean$risk_score <- predict(clean_cox_model, type = "lp")

# 2. Define the time horizon (same as your calibration plot)
time_horizon <- quantile(final_model_data_clean$exit_time, 0.80)

# 3. Calculate the Time-dependent ROC
roc_post_lasso <- timeROC(
  T = final_model_data_clean$exit_time,
  delta = final_model_data_clean$event,
  marker = final_model_data_clean$risk_score, 
  cause = 1,
  times = time_horizon,
  iid = TRUE # Keeps the data needed to calculate Confidence Intervals
)

# 4. Extract the Area Under the Curve (AUC) and Confidence Intervals
auc_value <- roc_post_lasso$AUC[2] # Index 2 corresponds to the time_horizon we passed
se_value <- roc_post_lasso$inference$vect_sd_1[2]

# Calculate 95% CI
ci_lower <- auc_value - 1.96 * se_value
ci_upper <- auc_value + 1.96 * se_value

cat("Time Horizon:", round(time_horizon, 2), "\n")
cat("AUC at Time Horizon:", round(auc_value, 3), "\n")
cat("95% CI: [", round(ci_lower, 3), "-", round(ci_upper, 3), "]\n\n")

# 5. Plot the ROC Curve
{plot(roc_post_lasso, 
     time = time_horizon, 
     col = "darkblue", 
     lwd = 2, 
     title = FALSE)

# Add a reference line and a title
abline(0, 1, lty = 2, col = "gray50")
title(main = paste("Post-LASSO Cox Model ROC\nat Time =", round(time_horizon, 1)))

# Add a legend with the AUC value
legend("bottomright", 
       legend = paste0("AUC = ", round(auc_value, 2), 
                       " (", round(ci_lower, 2), "-", round(ci_upper, 2), ")"),
       col = "darkblue", 
       lwd = 2, 
       bty = "n")}
```

## 2. rsf

### prepare the data by excluding the gene with frequency <= 1% (to reduce overfitting and increase interpretability)

```{r}
library(dplyr)
library(LTRCtrees) # For the random forest model
library(tidyr)

# 1. Define clinical columns to isolate the gene matrix
clinical_cols <- c(
  "X", "PATIENT_ID", "age_at_diagnosis",  "event", 
  "entry_time", "exit_time", "smoke_f", "stage_f", "prior_med_to_msk_f", 
  "prior_med_f", "gender"
)

# Isolate the true gene columns
gene_cols <- setdiff(colnames(nsclc), clinical_cols)
gene_matrix <- nsclc %>%
  select(all_of(gene_cols)) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.)))) 

# 2. Filter out genes with less than 1% frequency
N <- nrow(gene_matrix)
gene_frequencies <- colMeans(gene_matrix, na.rm = TRUE)
genes_to_keep_tfidf <- names(gene_frequencies[gene_frequencies >= 0.01])

# Subset matrix to kept genes
gene_matrix_filtered <- gene_matrix[, genes_to_keep_tfidf]

# 3. Calculate N and df for the filtered genes
df_filtered <- colSums(gene_matrix_filtered, na.rm = TRUE) 

# 4. Calculate the IDF Weights
idf_weights <- log(N / df_filtered)

# 5. Apply weights (TF * IDF)
gene_matrix_weighted <- sweep(
  x = gene_matrix_filtered, 
  MARGIN = 2, 
  STATS = idf_weights, 
  FUN = "*"
)

# 6. Recombine with clinical data and drop NAs for perfectly aligned complete cases
nsclc_tfidf_complete <- bind_cols(nsclc[, clinical_cols], gene_matrix_weighted) %>% 
  drop_na()

# Check your results
cat("Total patients (N complete):", nrow(nsclc_tfidf_complete), "\n")
cat("Original genes evaluated:", length(gene_cols), "\n")
cat("Genes kept for TF-IDF (>= 1% frequency):", length(genes_to_keep_tfidf), "\n")
```

### fit the rsf

```{r}
# 1. Define the features to include in the model
clinical_features <- c("age_at_diagnosis", "smoke_f", "stage_f", "prior_med_f", "gender")
all_features <- c(clinical_features, genes_to_keep_tfidf)

# 2. Dynamically construct the formula
rf_formula <- as.formula(
  paste("Surv(entry_time, exit_time, event) ~", paste(all_features, collapse = " + "))
)

# 3. Define mtry (rule of thumb is square root of the number of predictors)
m_try <- max(floor(sqrt(length(all_features))), 1)

# 4. Fit the Random Forest Model
set.seed(629)
rf_model_tfidf <- ltrcrrf(
  formula = rf_formula,
  data = nsclc_tfidf_complete,
  ntree = 500,        
  nodesize = 20,      
  mtry = m_try,       
  nsplit = 10         
)

# print(rf_model_tfidf)
```

### cross-validatoin (c-index)

```{r}
library(survival)

# Define the time horizon (80th percentile of follow-up time) for risk extraction
time_horizon_tfidf <- quantile(nsclc_tfidf_complete$exit_time, 0.80)

set.seed(629)
K <- 5
n <- nrow(nsclc_tfidf_complete)
fold_id <- sample(rep(1:K, length.out = n))
cv_cindex_rf_tfidf <- numeric(K)

for(k in 1:K){
  train_data <- nsclc_tfidf_complete[fold_id != k, ]
  test_data  <- nsclc_tfidf_complete[fold_id == k, ]
  
  # Fit the model on the training fold
  fit <- ltrcrrf(
    formula = rf_formula, 
    data = train_data, 
    ntree = 500, nodesize = 20, mtry = m_try, nsplit = 10
  )
  
  # Use predictProb to get survival at the exact time horizon
  prob_obj <- predictProb(
    fit,
    newdata = test_data,
    time.eval = time_horizon_tfidf
  )
  
  # Convert survival to risk score (1 - survival)
  survival_score <- as.numeric(prob_obj$survival.probs)
  test_data$risk_score <- 1 - survival_score
  
  # Calculate Concordance
  c_obj <- concordance(
    Surv(entry_time, exit_time, event) ~ risk_score,
    data = test_data,
    reverse = TRUE # TRUE because higher risk_score corresponds to shorter survival
  )
  
  cv_cindex_rf_tfidf[k] <- c_obj$concordance
}

cat("TF-IDF RF Mean C-index:", mean(cv_cindex_rf_tfidf, na.rm = TRUE), "\n")
cat("TF-IDF RF SD C-index:", sd(cv_cindex_rf_tfidf, na.rm = TRUE), "\n")
```

### calibration plot

```{r}
library(ggplot2)

# Get predictions on the full dataset at the time horizon
full_prob_obj <- predictProb(
  rf_model_tfidf,
  newdata = nsclc_tfidf_complete,
  time.eval = time_horizon_tfidf
)

# Extract predicted survival probability 
nsclc_tfidf_complete$pred_surv <- as.numeric(full_prob_obj$survival.probs)

# Create risk deciles based on predicted survival
nsclc_tfidf_complete$decile <- ntile(nsclc_tfidf_complete$pred_surv, 10)

# Calculate observed Kaplan-Meier estimates for each decile
calibration_data_rf_tfidf <- nsclc_tfidf_complete %>%
  group_by(decile) %>%
  group_modify(~ {
    
    km_fit <- survfit(Surv(entry_time, exit_time, event) ~ 1, data = .x)
    km_summary <- summary(km_fit, 
                          times = time_horizon_tfidf, 
                          extend = TRUE)
    
    data.frame(
      mean_pred = mean(.x$pred_surv, na.rm = TRUE),
      km_est = km_summary$surv
    )
  })

# Plot the calibration curve (using purple to distinctively mark the TF-IDF RF model)
ggplot(calibration_data_rf_tfidf, aes(x = mean_pred, y = km_est)) +
  geom_point(size = 3, color = "navy") + 
  geom_line(color = "navy", size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) + 
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = paste("Predicted Survival Probability at Time =", round(time_horizon_tfidf, 1)),
    y = "Observed Survival Probability (Kaplan-Meier)",
    title = "Calibration Plot: TF-IDF Weighted Random Forest",
    subtitle = "Red dashed line indicates perfect calibration"
  ) +
  theme_minimal()
```

### time-dependent ROC

```{r}
library(timeROC)

# Generate the risk scores from the calculated survival probabilities
nsclc_tfidf_complete$risk_score <- 1 - nsclc_tfidf_complete$pred_surv

# Calculate the Time-dependent ROC
roc_rf_tfidf <- timeROC(
  T = nsclc_tfidf_complete$exit_time,
  delta = nsclc_tfidf_complete$event,
  marker = nsclc_tfidf_complete$risk_score, 
  cause = 1,
  times = time_horizon_tfidf,
  iid = TRUE 
)

# Extract the Area Under the Curve (AUC) and 95% Confidence Intervals
auc_val_rf_tfidf <- roc_rf_tfidf$AUC[2] 
se_val_rf_tfidf <- roc_rf_tfidf$inference$vect_sd_1[2]

ci_lower_rf_tfidf <- auc_val_rf_tfidf - 1.96 * se_val_rf_tfidf
ci_upper_rf_tfidf <- auc_val_rf_tfidf + 1.96 * se_val_rf_tfidf

cat("TF-IDF RF - Time Horizon:", round(time_horizon_tfidf, 2), "\n")
cat("TF-IDF RF - AUC:", round(auc_val_rf_tfidf, 3), "\n")
cat("95% CI: [", round(ci_lower_rf_tfidf, 3), "-", round(ci_upper_rf_tfidf, 3), "]\n\n")

# Plot the ROC Curve
plot(roc_rf_tfidf, 
     time = time_horizon_tfidf, 
     col = "navy", 
     lwd = 2, 
     title = FALSE)

# Add reference line and title
abline(0, 1, lty = 2, col = "gray50")
title(main = paste("TF-IDF Random Forest ROC\nat Time =", round(time_horizon_tfidf, 1)))

# Add legend with the final AUC value
legend("bottomright", 
       legend = paste0("AUC = ", round(auc_val_rf_tfidf, 2), 
                       " (", round(ci_lower_rf_tfidf, 2), "-", round(ci_upper_rf_tfidf, 2), ")"),
       col = "navy", 
       lwd = 2, 
       bty = "n")
```


